\textbf{Estimating the proportion of beneficial mutations that are not adaptive in mammals}

### 2024-12-27 manuscript/previous-versions/supp-mat-PLOS-revision.tex
Moved tex files into a single folder

    In the figures below, the first line is the name of the gene, the second line the position of the mutation in the OrthoMam protein-coding DNA alignment, and the arrow in the third line corresponds to the nucleotide position of the mutation.
    Finally, for reproducibility, we generated the fasta alignment for all $\SphyBen$ mutations that we detected, either in the terminal lineage or in segregating polymorphism.
    These fasta files are available at Zenodo (\url{https://doi.org/10.5281/zenodo.7878953}), across all populations, under the compressed archive \textit{alignment\_focused\_around\_non\_adaptive\_mutations.zip} and we provide a python script to obtain the figures shown below given these alignments.

@@

    \section{Correlation with diversity}

@@

    \newpage
    \section{Excluding genes under adaptation}

@@

    \newpage
    \section{Including divergence data for the estimation of $\Spop$}

    \subsection{Including divergence in polyDFE}
    Divergence data (number of substitutions per site) can also be included into polyDFE to estimate the DFE.
    For the class of a given class of selection coefficient ($\Sphyclass \in \{\SphyDel, \SphyNeu, \SphyBen \}$), the number of substitutions has already been computed and is given by $D\left( \Sphyclass \right)$ (see eq.~6), and the number of sites is given by $L \left( \Sphyclass \right)$ (see eq.~4).
    Otherwise, the procedure is the same as described section \textit{Scaled selection coefficients ($\Spop$) in a population-based method}.

    \subsection{Probability for beneficial mutations to be non-adaptive}
    \begin{center}
        \captionof{table}{$\proba[\SphyBen\given \SpopBen ]$ when including divergence to estimate $\Spop$.}
        \scriptsize
        \begin{longtable*}{|l|l|r|r|r|r|r|r|}
            \toprule
            Population & Species & $\Ne$ & $\proba[\SphyBen]$ & $\proba [ \SpopBen ]$ & $\frac{\proba[\SphyBen]}{\proba[ \SpopBen ]}$ & $\proba [ \SpopBen \given \SphyBen]$ & $\proba[\SphyBen\given \SpopBen ]$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{8}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

@@

    \section{Discrete distribution of $\Spop$ at the population scale}

    \subsection{polyDFE model D (discrete distribution)}
    Additionally to including divergence, with also tested our prediction with polyDFE model D instead of model C.
    In polyDFE model D, the DFE of non-synonymous mutations is given as a discrete DFE of $K$ categories (instead of a continuous distribution in model C), where the selection coefficients of each category $i$ ($1 \leq i \leq K$) are fixed parameters $\Spop_i$, and each value $\Spop_i$ has a probability $p_i$ (estimated), with $\sum_{i=1}^{K} p_i =1$.
    We used $K=6$ with $\Spop_1 = -500$, $\Spop_2 = -4$, $\Spop_3 =-1$, $\Spop_4 = 0$, $\Spop_5 = 1$, $\Spop_6 = 4$.\\
    For each class of selection $\Sphyclass$, the parameters $p_i$ ($i \in \{ 1 \leq i \leq 6 \}$) were used to compute $\proba [ \SpopDel \given  \Sphyclass] $, $\proba [ \SpopNeu \given \Sphyclass]$, and $\proba [ \SpopBen \given \Sphyclass]$ as:
    \begin{align}
        \proba [ \SpopDel \given  \Sphyclass] &= \proba [ \Spop < -1 \given \Sphyclass ] = p_1 + p_2 \label{eq:polyProbaDel-mD} \\
        \proba [ \SpopNeu \given \Sphyclass] &= \proba [ -1 < \Spop < 1 \given \Sphyclass ] = p_3 + p_4 + p_5,  \\
        \proba [ \SpopBen \given \Sphyclass] &= \proba [  \Spop > 1 \given \Sphyclass ] = p_6.  \label{eq:polyProbaAdv-mD}
    \end{align}

@@

    Altogether, comparing this table other estimations, we acknowledge that the exact proportion of non-adaptive among all beneficial ones is dependent on the model used to estimate the DFE.
    However, we can still see that non-adaptive beneficial mutations are positively selected compared to neutral and deleterious ones, and this result is not sensitive to the underlying DFE at the population scale.

>>>>

    In the figures below, the first line is the name of the gene, the second line the position of the mutation in the OrthoMam protein-coding DNA alignment, and the arrow in the third line corresponds to the nucleotide position of the mutation.
    Finally, for reproducibility, we generated the fasta alignment for all $\SphyBen$ mutations that we detected, either in the terminal lineage or in segregating polymorphism.
    These fasta files are available at Zenodo (\url{https://doi.org/10.5281/zenodo.7878953}), across all populations, under a zip file (\textit{alignment\_around\_non\_adaptive\_mutations.zip}) and we provide a python script to obtain the figures shown below given these alignments (\textit{plot\_variations.py}).

@@

    \newpage
    \section{Correlation with diversity}

@@

    \newpage
    \subsection{Simulations of site-frequency spectra}

    We performed simulations based on population-genetic formalism to test whether our approach is able to recover the proportion of beneficial mutations that are non-adaptive in synthetic polymorphism datasets.
    More specifically, we performed simulations of site-frequency spectra, thus at the population-genetic scale, by using a mixture of the DFEs that is predicted from the phylogenetic mutation-selection model, and adding adaptive mutations to this DFE.
    In practice, simulations are performed under the Poisson Random Field (PRF) framework of population genetics\cite{sawyer_population_1992}.
    Each simulation generates a synthetic synonymous SFS, and a synthetic non-synonymous SFS for the mutations in the category $x \in \{ \SphyDel, \SphyNeu, \SphyBen \}$.

    Technically, for non-synonymous mutations, we first draw a random sample of the selection coefficients from the mixture of DFE using as parameter $\proba [y \given x]$ for $y \in \{ \SpopDel, \SpopNeu, \SpopBen \}$.
    Importantly, $\proba [\SpopBen | \SphyDel]$ and $\proba [\SpopBen | \SphyNeu]$ are considered adaptive mutations, and thus their selection coefficients have a selection coefficient of $S_{\text{adaptive}}=10$.
    All the others are drawn for the DFEs that are predicted from the phylogenetic mutation-selection model.
    Synonymous mutations are considered neutral and have $S=0$.

    Given these selection coefficients, to obtain the expected SFS we integrate over the limiting density of frequency that mutations are segregating in the population, as described from eqs. 1--2 in \textcite{tataru_inference_2017}.
    And finally, given this expected SFS, we perform a Poisson random sampling to obtain a sampled SFS of size n matching the total number of polymorphism to simulate.

    In summary, the simulations generates a synonymous SFS, and a non-synonymous SFS for the mutations in the category $x \in \{ \SphyDel, \SphyNeu, \SphyBen \}$, and require as input the following:
    \begin{itemize}
        \item Total number of polymorphisms to simulate (synonymous and non-synonymous).
        \item The DFE of $\Sphy$ predicted from the phylogenetic mutation-selection.
        \item Estimates of $\proba [y \given x]$ for $y \in \{ \SpopDel, \SpopNeu, \SpopBen \}$.
        \item Selection coefficient for adaptive mutations, here $S_{\text{adaptive}}=20$.
        \item The sampled SFS size $n$, here $n=16$ (i.e.\ twice the number of sampled individuals).
    \end{itemize}

    From these two SFS (synonymous and non-synonymous), we run polyDFE to obtain $\proba [\SpopDel \given x]$, $\proba [\SpopNeu \given x]$ and $\proba [\SpopBen \given x]$ (eqs.\ 9--11 in the main manuscript).
    By doing so for the 3 categories $x \in \{ \SphyDel, \SphyNeu, \SphyBen \}$, we can ultimately compute $\proba [\SphyBen \given \SpopBen]$ (eqs.\ 14--15 in the main manuscript), which can be then be compared to the ground truth.

    \newpage
    \begin{center}
        \captionof{figure}{Simulations of site-frequency spectra.}
        \includegraphics[width=\linewidth, page=1]{artworks/simulations.pos.pdf}
    \end{center}

    We used 8 different set of parameters ($\proba [y \given x]$ and numbers of polymorphism) and simulated 10 replicates for each set of parameters.
    These set of input parameters are provided to match $\proba [\SphyBen \given \SpopBen]$ of the following populations: Set 1 for Admixed American (Homo sapiens), Set 2 for African (Homo sapiens), Set 3 for Nevis (Chlorocebus sabaeus), Set 4 for Iran (Ovis orientalis), Set 5 for Uganda (Bos taurus), Set 6 for Morocco (Ovis aries), Set 7 for Equus caballus, Set 8 for Gambia (Chlorocebus sabaeus).
    For each synthetic data we run polyDFE and derived $\proba [\SphyBen \given \SpopBen]$ which is then compared to the ground truth (calculated analytically from the set parameters $\proba [y \given x], x \in \{ \SphyDel, \SphyNeu, \SphyBen \}, y \in \{ \SpopDel, \SpopNeu, \SpopBen \}$).
    These simulations show that we are able to retrieve accurately the proportion of non-adaptive mutations in those synthetic datasets.

@@

    \section{Excluding genes under adaptation}

    \section{Including substitutions in the terminal lineage to estimate $\Spop$}

    \captionof{figure}{Including substitutions in the terminal lineage to estimate $\Spop$.}
    \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba.div.pdf}
    Same as figure 2 in the main text, but including substitutions in the terminal lineage to estimate $\Spop$.
    At the phylogenetic scale (A), we estimated the amino-acid fitness for each site from protein-coding DNA alignments using mutation-selection codon models.
    For every possible mutation, the difference in amino-acid fitness before and after the mutation allows us to compute the selection coefficient at the phylogenetic scale ($\Sphy$).
    Depending on $\Sphy$ (B) mutations can be predicted as deleterious ($\SphyDel$), nearly-neutral ($\SphyNeu$) or beneficial non-adaptive mutations ($\SphyBen$) toward a fitter amino acid and repairing existing functions.
    For each population, each observed single nucleotide polymorphism (SNP) segregating can also be classified according to its $\Sphy$ value (C).
    Within the terminal lineage leading to this population, each substitution can be classified according to its $\Spop$ value (C).
    Non-synonymous polymorphisms and divergence, contrasted to synonymous polymorphisms and divergence (deemed neutral) is used to estimate selection coefficients (D-E) at the population scale ($\Spop$), for each class of selection ($\SphyDel$, $\SphyNeu$, $\SphyBen$).
    We can thus assess whether $\Sphy$ predicts $\Spop$ and compute \textit{precision} (F) and \textit{recall} (G) for each class.
    The \textit{recall} value for class $\SphyBen$ is the probability for beneficial mutations to be non-adaptive (G).

    \subsection{Including substitutions in polyDFE}
    Divergence data (number of substitutions per site) can also be included into polyDFE to estimate the DFE.
    For the class of a given class of selection coefficient ($\Sphyclass \in \{\SphyDel, \SphyNeu, \SphyBen \}$), the number of substitutions has already been computed and is given by $D\left( \Sphyclass \right)$ (see eq.~6), and the number of sites is given by $L \left( \Sphyclass \right)$ (see eq.~4).
    Otherwise, the procedure is the same as described section \textit{Scaled selection coefficients ($\Spop$) in a population-based method}.

    \newpage
    \subsection{Probability for beneficial mutations to be non-adaptive}
    \begin{center}
        \captionof{table}{$\proba[\SphyBen\given \SpopBen ]$ when including divergence to estimate $\Spop$.}
        \scriptsize
        \begin{longtable*}{|l|l|r|r|r|r|r|r|}
            \toprule
            Population & Species & $\Ne$ & $\proba[\SphyBen]$ & $\proba [ \SpopBen ]$ & $\frac{\proba[\SphyBen]}{\proba[ \SpopBen ]}$ & $\proba [ \SpopBen \given \SphyBen]$ & $\proba[\SphyBen\given \SpopBen ]$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{8}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

    \newpage
    \subsection{Fitting one or three distributions of fitness effects}

    We also performed another test to assess that fitting the same functional form of DFE (distribution of fitness effects) to three categories of selection (i.e. $\SphyDel, \SphyNeu, \SphyBen$) is a sound methodology.
    Typically, $\proba [ \SpopBen ]$ can be computed from two different estimates, by either splitting data (polymorphism within populations and substitutions in the terminal lineages) in three category (i.e. $\SphyDel, \SphyNeu, \SphyBen$) or not at all.
    In the case of splitting data in three categories and fitting three DFE, $\proba [ \SpopBen ]$ is obtained by summing over the three category using the law of total probability (eq.\ 15 in the main manuscript):

    \begin{equation}
        \proba [ \SpopBen ] \text{ (three DFE) } = \proba [\SpopBen \given \SphyDel ] \times \proba [\SphyDel ] + \proba [\SpopBen \given \SphyNeu ] \times \proba [\SphyNeu ] + \proba [\SpopBen \given \SphyBen ] \times \proba [\SphyBen ].
        \label{eq:total_proba}
    \end{equation}

    But it is also possible to compute $\proba [ \SpopBen ]$ directly from fitting one DFE to the full data (polymorphism within populations and substitutions in the terminal lineages), not splitting at all, and then $\proba [ \SpopBen ]$ is obtained from the parameters of the DFE (eq.\ 11 in the main manuscript):
    \begin{align}
        \proba [ \SpopBen ] \text{ (one DFE) } &= \proba [ \Spop > 1 ] = p_b \int_{1}^{+\infty} f_{e}(\Spop; \AdvMean) \der \Spop. \label{eq:polyProbaAdv}
    \end{align}

    We reproduced this experiment for $\proba [ \SpopDel ]$, $\proba [ \SpopNeu ]$ and $\proba [ \SpopBen ]$ as shown below.
    The goodness of fit is calculated as the coefficient of determination ($R^2$) for the prediction $\proba [ \SpopBen ] \text{ (three DFE) } = \proba [ \SpopBen ] \text{ (one DFE) }$, meaning calculating the residuals of the fit to the 1:1 line.

    \newpage
    \begin{center}
        \captionof{figure}{$\proba [ \SpopDel ]$ when fitting a single or three DFEs.}
        \includegraphics[width=0.75\linewidth, page=1]{artworks/results.reg.neg.div.scatter.pdf}
    \end{center}
    \begin{center}
        \captionof{figure}{$\proba [ \SpopNeu ]$ when fitting a single or three DFEs.}
        \includegraphics[width=0.75\linewidth, page=1]{artworks/results.reg.weak.div.scatter.pdf}
    \end{center}
    \begin{center}
        \captionof{figure}{$\proba [ \SpopBen ]$ when fitting a single or three DFEs.}
        \includegraphics[width=0.75\linewidth, page=1]{artworks/results.reg.pos.div.scatter.pdf}
    \end{center}

@@

    \section{Non-parametric distribution of fitness effects}

    \subsection{polyDFE model D (non-parametric)}
    Additionally to including divergence, with also tested our prediction with polyDFE model D instead of model C.
    In polyDFE model D, the DFE of non-synonymous mutations is given as a discrete DFE of $K$ categories (instead of a continuous distribution in model C), where the selection coefficients of each category $i$ ($1 \leq i \leq K$) are fixed parameters $\Spop_i$, and each value $\Spop_i$ has a probability $p_i$ (estimated), with $\sum_{i=1}^{K} p_i =1$.
    We used $K=6$ with $\Spop_1 = -500$, $\Spop_2 = -4$, $\Spop_3 =-1$, $\Spop_4 = 0$, $\Spop_5 = 1$, $\Spop_6 = 4$.\\
    For each class of selection $\Sphyclass$, the parameters $p_i$ ($i \in \{ 1 \leq i \leq 6 \}$) were used to compute $\proba [ \SpopDel \given  \Sphyclass] $, $\proba [ \SpopNeu \given \Sphyclass]$, and $\proba [ \SpopBen \given \Sphyclass]$ as:
    \begin{align}
        \proba [ \SpopDel \given  \Sphyclass] &= \proba [ \Spop < -1 \given \Sphyclass ] = p_1 + p_2 \label{eq:polyProbaDel-mD} \\
        \proba [ \SpopNeu \given \Sphyclass] &= \proba [ -1 < \Spop < 1 \given \Sphyclass ] = p_3 + p_4 + p_5,  \\
        \proba [ \SpopBen \given \Sphyclass] &= \proba [  \Spop > 1 \given \Sphyclass ] = p_6.  \label{eq:polyProbaAdv-mD}
    \end{align}

@@

    Altogether, comparing this table other estimations, we acknowledge that the exact proportion of non-adaptive among all beneficial ones is dependent on the model used to estimate the DFE.
    However, we can still see that non-adaptive beneficial mutations are positively selected compared to neutral and deleterious ones, and this result is valid for different underlying DFE.
